Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-13 paper/main.tex
some formatting in the extended version
more examples and technical remarks

With two concrete examples from open-source projects we demonstrate
that the compositionality of our method is indeed crucial in practice.
\jan{what projects?}

Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations of block-based
cryptographic primitives: Data of arbitrary length, is consumed in
blocks and the leftover is stored in a buffer for future use when more
data is available.  This looping pattern is particularly well handled
by our method. If $N \geq 8$, $\toolname$ infers the bound
$\frac{N}{8} |[0,l]|$, but if $N<8$ $\toolname$ infers $7\frac{8-N}{8}
+ \frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the
cost in the second loop (\code{tick(1)}) are random choices and
$\toolname$ would also derive tight bound for other values.

@@

There exist quantitative logics that are integrated into separation
logic~\cite{Atkey10,HoffmannMZ13} and they are closely related to our
quantitative logic.  However, the purpose of these logics is slightly
different since they focus on the verification of bounds that depend
on the shape of heap data structures and they are not implemented for
C.  Also closely related to our logic is a VDM-style logic for
reasoning about resource usage of an abstract fragment of JVM byte
code by Aspinall et al.~\cite{AspinallBHLM07}.  Their logic is not
Hoare-style, does not target C code, and is not designed for
interactive bound development but to produce certificates for bounds
derived for high-level functional programs.
}{}
\begin{table}[t]
\centering
\begin{tabular}{r|c|c|c}
Function & LoC & Bound & Time (s) \\
\hline
\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
\end{tabular}
\vspace{.1cm}
\caption{Derived bounds for functions from cBench.}
\label{tab:cbench}
\end{table}

@@

We have demonstrated that our approach improves the state-of-the-art
in resource bound analysis for C programs in three ways. First, our
technique is naturally compositional, tracks size changes of
variables, and can abstractly specify the resource cost of functions
(\pref{sec:AAA}). Second, it is easily combinable with established
qualitative verification to guide semi-automatic bound derivation
(\pref{sec:anno}). Third, we have shown that the local inference rules
of the derivation system automatically produce easily checkable
certificates for the derived bounds (\pref{sec:soundness}).

Our system is the first amortized resource analysis for C programs.
It addresses the longstanding open problem of extending automatic
amortized resource analysis to compute bounds for programs that loop
on singed integers and to deal with non-linear control flow that is
introduced by break and return statements.

@@

@@

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcomes of a program
statement.  We define it as follows.
\begin{align*}
\safeK n B R Q \cont & := & & \safe n B {\code{break}} \cont \\
& & \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
& & \land& \safe n Q {\code{skip}} \cont
\end{align*}
%
We can now define the \emph{semantic validity} of a judgement $B; R
\vdash \htriple P S Q$ of the quantitative logic without function
context as $\valid n B R P S Q :=$
\begin{align*}
& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B{+}M_b{+}x} {R{+}x} {Q{+}x} \cont) \\
& \implies \safe m {P+x} S \cont \; .
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic. This refinement is necessary to have a stronger
induction hypothesis available during the proof.
%
We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
S Q$ implies $\valid n B R P S Q$.

Using the semantic validity of triples we define the validity
of a function context $\Delta$, written $\validC n \Delta$, as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$. A full
judgement that mentions a non-empty function context
$\Delta$ is in fact a guarded statement: it makes
assumptions on some functions' behavior.  The predicate
$\validC n \Delta$ gives the precise meaning of the
assumptions made.  It is also step-indexed to prove the
soundness of the {\sc L:Extend} rule by induction.
%
We are now able to state the soundness of the quantitative logic.
%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
  $
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $
\end{theorem}
%
\noindent
The difference $\delta = 1$ between the index in the triple
validity and the one in context validity arises from
the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
{\sc L:Call}, the language semantics makes one step and
proceeds with the function body, so we must have
$\delta \le 1$ to use the assumptions in $\Delta$.
For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
is a valid context for $n$ steps.  The induction hypothesis
in that case says that if $\Delta \cup \Delta'$ is valid
for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
So if we want to solve this goal by induction, it is
necessary that $\delta \ge 1$.  These two constraints force
$\delta$ to be exactly one in the theorem statement.

@@

\begin{figure}[t!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:RetSeq}
{}
{ (\state, \code{return}~x, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{}
{ (\state, \code{return}~x, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_r)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Call}
{ \fenv f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep{M}
(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep{M}
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{loop}~S, \cont, \cost {-} M_l)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma
\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \cont, \cost')
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma
\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_2, \cont, \cost')
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, S, \cont, \cost {-} M_s)
}
%
\and
\Rule{S:Tick}
{ }
{ (\state, \code{tick}(n), \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_t(n))
}

\end{mathpar}
\caption{Rules of the operational semantics of statements.}
\label{fig:opsem}
\end{figure}

>>>>

With two concrete examples from open-source projects  we demonstrate
that the compositionality of our method is indeed crucial in practice.

Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations
of block-based cryptographic primitives: Data of arbitrary length, is
consumed in blocks and the leftover is stored in a buffer for future use
when more data is available.  It is present in all the block encryption
routines of PGP.  The same pattern is also used in performance critical
code to unroll a loop, for example we found it in a bit manipulating
function of libtiff library and a CRC computation routine of MAD,
an MPEG audio decoder.  This looping pattern is particularly well
handled by our method. If $N \geq 8$, $\toolname$ infers the bound
$\frac{N}{8} |[0,l]|$, but if $N<8$ $\toolname$ infers $7\frac{8-N}{8} +
\frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the cost
in the second loop (\code{tick(1)}) are random choices and $\toolname$
would also derive tight bound for other values.

@@

There exist quantitative logics that are integrated into separation
logic~\cite{Atkey10,HoffmannMZ13} and they are closely related to our
quantitative logic.  However, the purpose of these logics is slightly
different since they focus on the verification of bounds that depend
on the shape of heap data structures and they are not implemented for
C.  Also closely related to our logic is a VDM-style logic for
reasoning about resource usage of an abstract fragment of JVM byte
code by Aspinall et al.~\cite{AspinallBHLM07}.  Their logic is not
Hoare-style, does not target C code, and is not designed for
interactive bound development but to produce certificates for bounds
derived for high-level functional programs.
}{}

\begin{table}[t]
\centering
\begin{tabular}{r|c|c|c}
Function & LoC & Bound & Time (s) \\
\hline
\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
\code{MD5Final} & 195 & 141 & 0.22 \\
\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
\code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
\hline
\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1
\end{tabular}
\vspace{.1cm}
\caption{Derived bounds for functions from cBench. \jan{FUCKING backedge metric}}
\label{tab:cbench}
\end{table}

@@

We have demonstrated that our approach improves the state-of-the-art
in resource bound analysis for C programs in three ways. First, our
technique is naturally compositional, tracks size changes of
variables, and can abstractly specify the resource cost of functions
(\pref{sec:AAA}). Second, it is easily combinable with established
qualitative verification to guide semi-automatic bound derivation
(\pref{sec:anno}). Third, we have shown that the local inference rules
of the derivation system automatically produce easily checkable
certificates for the derived bounds (\pref{sec:soundness}).
%
Our system is the first amortized resource analysis for C programs.
It addresses the longstanding open problem of extending automatic
amortized resource analysis to compute bounds for programs that loop
on singed integers and to deal with non-linear control flow that is
introduced by break and return statements.

@@

\begin{figure*}[t!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:RetSeq}
{}
{ (\state, \code{return}~x, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{}
{ (\state, \code{return}~x, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_r)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Call}
{ \fenv f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep{M}
(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep{M}
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{loop}~S, \cont, \cost {-} M_l)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma
\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \cont, \cost')
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma
\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_2, \cont, \cost')
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, S, \cont, \cost {-} M_s)
}
%
\and
\Rule{S:Tick}
{ }
{ (\state, \code{tick}(n), \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_t(n))
}

\end{mathpar}
\caption{Rules of the operational semantics of statements.}
\label{fig:opsem}
\end{figure*}

@@

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcomes of a program
statement.  We define it as follows.
\begin{align*}
\safeK n B R Q \cont & :=  \safe n B {\code{break}} \cont \\
 \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
 \land& \safe n Q {\code{skip}} \cont
\end{align*}
%
We can now define the \emph{semantic validity} of a judgement $B; R
\vdash \htriple P S Q$ of the quantitative logic without function
context as $\valid n B R P S Q :=$
\begin{align*}
& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B{+}M_b{+}x} {R{+}x} {Q{+}x} \cont) \\
& \implies \safe m {P+x} S \cont \; .
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic. This refinement is necessary to have a stronger
induction hypothesis available during the proof.
%
We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
S Q$ implies $\valid n B R P S Q$.

Using the semantic validity of triples we define the validity
of a function context $\Delta$, written $\validC n \Delta$, as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$. A full
judgement that mentions a non-empty function context
$\Delta$ is in fact a guarded statement: it makes
assumptions on some functions' behavior.  The predicate
$\validC n \Delta$ gives the precise meaning of the
assumptions made.  It is also step-indexed to prove the
soundness of the {\sc L:Extend} rule by induction.
%
We are now able to state the soundness of the quantitative logic.
%
\begin{theorem}[Soundness]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
  $
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $
\end{theorem}
%
\noindent
The difference $\delta = 1$ between the index in the triple
validity and the one in context validity arises from
the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
{\sc L:Call}, the language semantics makes one step and
proceeds with the function body, so we must have
$\delta \le 1$ to use the assumptions in $\Delta$.
For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
is a valid context for $n$ steps.  The induction hypothesis
in that case says that if $\Delta \cup \Delta'$ is valid
for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
So if we want to solve this goal by induction, it is
necessary that $\delta \ge 1$.  These two constraints force
$\delta$ to be exactly one in the theorem statement.

@@
